---
title: "HW#4"
author: "Natalia"
date: "13 April 2019"
output: pdf_document
---
```{r}
set.seed(42)

setwd("C:\\Users\\Natalia\\Desktop\\ITMO\\R\\R_task#4")
raw_df = readRDS("weather.rds")

```

#Exploring dataset
```{r}
summary(raw_df)
head(raw_df)
str(raw_df)
View(raw_df)

```

# Problems of the dataset:

1) "X" collumn is just an index which is not needed for further analysis - have to be deleted.
2) It is better to assign our samples as [Year+month+day] from X number of columns.
3) Split [measure] column into multiple columns


```{r}
# Packages included
library(tidyr)
library(stringr)
# Get rid of [X] column
tidy_df = raw_df[,-1]
head(tidy_df)
# Make dataframe from wide to long
tidy_df = gather(tidy_df, day, value, X1:X31)
# Make dataframe from long to wide
tidy_df = spread(tidy_df, measure, value)
# Assign [Year+month+day] as samples names
tidy_df$day = str_sub(tidy_df$day, 2, 3)
tidy_df = unite(tidy_df, Date, year, month, day, sep = '_')
head(tidy_df)
```

3) Then it was noticed that some samples have all NAs data because of unexisting days (February 30, etc). 
   Also, the measurements in the end of the table are NAs as well.  


```{r}
tidy_df = na.omit(tidy_df)
```

4) Formatting of the columns:

```{r}
tidy_df$PrecipitationIn = as.numeric(tidy_df$PrecipitationIn, na.string='T')
# Make columns with values as numbers numeric
tidy_df[,c(2,4:23)] = lapply(tidy_df[,c(2,4:23)], as.numeric)
# Check the dataframe after formatting
row.names(tidy_df) = 1:nrow(tidy_df)
summary(tidy_df)
```

```{r}
# Make [Events] column  as factor
tidy_df$Events = as.factor(tidy_df$Events)
levels(tidy_df$Events)[1] = "Ordinary day"
head(tidy_df)
```

5) Plot the numeric columns data

```{r}
boxplot(tidy_df[, c(2, 3:23)])
```

6) Obvious outlier was detected - the [Humidity] was 1000% at one point of measurements that seems to be unreal   data (extra "0" was added mistakenly) and should be changed for more realistic value = 100%.
 
```{r}
tidy_df[135,6] = 100
# Check the plot after additing
boxplot(tidy_df[, c(2, 3:23)])
```

# Finally, the dataset seems to be clear and tidy after processing and ready for further analysis.

Difference between "raw" and "tidy" dataframes.
 
```{r}
head(raw_df)
head(tidy_df)
```

